\documentclass[acmsmall,screen,review]{acmart}

\usepackage{proof}
\usepackage{fancyvrb}
\usepackage{minted}
\setminted{fontsize=\small} % Remove this for larger code snippets

\usepackage{tikz}
\usepackage{graphicx}
\usepackage{mathpartir}
\usepackage{color}
\usepackage{pgflibraryarrows}
\usepackage{pgflibraryshapes}
\usepackage{pgfbaseimage}
\usepackage{pgfplots}
\usepackage{hyperref}

\usetikzlibrary{decorations.text}
\usetikzlibrary{trees}
\usetikzlibrary{fadings}
\usetikzlibrary{arrows.meta}

\usepackage{xcolor}

\usepackage{amsthm}

\let\Bbbk\relax
\usepackage{amssymb}

\let\Bbbk\relax
\usepackage{amssymb}

\usepackage[ruled,vlined]{algorithm2e}
\SetAlgoLined
\SetAlgoNoEnd
\SetArgSty{textup}
\SetKwProg{Program}{Program}{\hspace{0.3em}:}{}
\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
\SetKwIF{Match}{ElseMatch}{Otherwise}{match}{with}{else}{else}{endmatch}

\usepackage{fancyhdr}

\usepackage{booktabs}

\usepackage{multicol}

\setcopyright{acmlicensed}
\copyrightyear{2026}
\acmYear{2026}
\acmDOI{XXXXXXX.XXXXXXX}

\acmJournal{JACM}
\acmVolume{37}
\acmNumber{4}
\acmArticle{111}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}


%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

\include{definitions}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

\title{There Must Be Some Kind of Mistake: Distinguishing Errors from Uncertainty in Type Checking}


%% \author{Aaron Stump}
%% \email{stumpaa@bc.edu}
%% \affiliation{%
%%   \institution{Boston College}
%%   \city{Boston}
%%   \state{Massachusetts}
%%   \country{USA}
%% }


%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{\ }%Blanchette et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A recent paper by Garby, Bahr, and Hutton proposes a relational
  specification for type checkers, where types computed statically are
  related to the types of computational results.  The relation in
  question makes type errors less than non-error types.  This paper
  refines this relation by introducing explicit types for terms which
  are guaranteed to exhibit runtime errors of various kinds.  These
  error types are distinct from a type representing uncertainty about
  the term's runtime behavior.  The two kinds of errors considered are
  finite failure and divergence.  To manage the complexity induced by
  the richer partial order on types, the paper uses a monadic type
  checker, which is then related to a fuel-bounded monadic
  interpreter.  Proof complexity is also managed monadically, using a
  binary indexed monad to relate typing and evaluation.  The
  verification is carried out in Agda.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011074.10011099.10011692</concept_id>
       <concept_desc>Software and its engineering~Formal software verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10010124.10010138.10010142</concept_id>
       <concept_desc>Theory of computation~Program verification</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Formal software verification}
\ccsdesc[300]{Theory of computation~Program verification}
\ccsdesc[500]{Theory of computation~Type theory}

\keywords{type checking, nontermination, type errors, monads}

%\received{20 February 2007}
%\received[revised]{12 March 2009}
%\received[accepted]{5 June 2009}

\theoremstyle{definition}
\newtheorem{specification}[section]{Specification}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

\begin{figure}
\begin{verbatim}
data Value = I Int | B Bool | Error
data Expr = Val Value | Add Expr Expr | If Expr Expr Expr
data Type = INT | BOOL | ERROR

tval :: Value -> Type
texp :: Expr -> Type
eval :: Expr -> Value

instance Ord Type where
 t ⩽ t'  =  t == ERROR  ||  t == t'
\end{verbatim}
\caption{Definitions from \citet{garby+25}}
\label{fig:orig}
\end{figure}


\citet{garby+25} present the derivation of a type checker for a simple
expression language, in the style of program calculation.  With this
approach, one tries to prove a specification about code that has yet
to be determined, and discovers the definitions (or constraints on
them) necessary in the course of attempting to complete the proof.
Their specification states that the type computed by the type checker
\verb|texp| for an \textit{Expr} $e$ is an approximation to the true
type of $e$.  The true type is the type computed (by a function
\textit{tval}) for the value of $e$, as produced by an evaluator
\textit{eval}.  The approximation \verb|(⩽)| states that the
\textit{ERROR} type returned by the type checker to signal a type
error approximates any of the types for values.  With the types shown
in Figure~\ref{fig:orig} (following the authors in writing Haskell
with unicode), their specification is
\[
\textit{texp}\ e \le \textit{tval}\ (\textit{eval}\ e)
\]
\noindent While acknowledging that usually one would implement the
evaluator monadically to handle runtime errors, they opt to use
simpler typing, and include
\verb|Error| as a \verb|Value|.

In this paper, we build on the prior work of Garby et al. in the following ways:

\begin{enumerate}
\item We extend the ordering on types to distinguish known errors from uncertainty.  So we now have
  types \verb|UNKNOWN| and \verb|FAIL|.  If an expression is of type \verb|FAIL|, that means that
  it is guaranteed to produce a runtime error.  \verb|UNKNOWN| represents a term whose runtime behavior
  cannot be assessed by the type checker.
  
  \item We add a form of possibly diverging computation to the simple
    expression language, via a form of unbounded search.  This leads
    to richer typing: instead of just \verb|BOOL| and \verb|INT| for
    terms producing values of those types, we now distinguish between
    terms which are known to terminate or ones which might diverge.  So we
    have types \verb|BOOL b| and \verb|INT b|, where boolean \verb|b|
    indicates whether possible divergence
    (\verb|True|) or guaranteed termination (\verb|False|).  The
    \verb|FAIL| type is updated similarly, to distinguish terms that
    are guaranteed to fail at runtime, from those that fail or else diverge.
    For computations that are \emph{guaranteed} to diverge, we
    add a type \verb|LOOP|.
    
  \item With this more complex ordering on types, we adopt monadic definitions of the type
    checker and the evaluator.

  \item Using these enriched definitions, we prove the specification
    proposed by~\citet{garby+25}, in Agda.  Here, this specification
    involves relating two monadic computations, namely the type
    checker and the evaluator.  To do this concisely, we make use of a
    binary indexed monad in the proof.

    \end{enumerate}

The motivation for this development is to enrich the kind of
information the type checker can provide.  Usually (and
in~\citet{garby+25}), type checkers do not distinguish between
situations that might lead to runtime errors, and those which must.
Such a distinction would pave the way to providing new error
information to programmers, when their programs do not pass the type
checker.  One might wish, for example, to report example inputs to a
function which are guaranteed to lead to a runtime error.  It is
common practice to provide such information in the world of model
checking, in the form of error traces~\cite{jhala+09}.  Formalizing
such error reporting remains, however, for future work.  We here just
address the issue of relating typing and evaluation with a more
refined notion of untypability.

All code following is written in Agda (checked with version 2.7.0.1),
making modest use of an
\href{https://github.com/cedille/ial}{alternative standard library}
for basic data structures.  Readers unfamiliar with Agda will find
that our code looks very similar to Haskell, except for the use of
unicode and user-defined mixfix syntax, where underscores in the names
of functions indicate argument positions.  Here is a brief summary of
other Agda features we use. Pattern-matching equations can be extended
to match on an additional expression using \verb|with|.  The
additional argument is listed following a \verb;|; in subsequent
equations.  The library provides types $\mathbb{Z}$ of integers,
$\mathbb{N}$ of unary natural numbers, and $\mathbb{B}$ of booleans,
with values \verb|tt| and \verb|ff|.  Unlike Haskell, Agda allows the
same name to be used for a constructor of two different types, but the
same name may not be used at both the term and type levels (because
the notion of level is more complex in Agda).  Finally, inputs to
functions can be designated implicit using curly braces.  Agda tries
to infer values for those at call sites.  Occasionally this fails, and
then the programmer may resort to writing those arguments explicitly,
in curly braces.

\section{The Expression Language}

The syntax for values and expressions is in Figure~\ref{fig:syn}.  
Similarly to Garby et al., we work with a first-order language with
integer and boolean values.  Our expression syntax includes values (\verb|Value|),
additions (\verb|Add|), and conditionals (\verb|Cond|), just like
Garby et al.  We add a test for zero (\verb|IsZero|), to have a way to
compute a boolean.  Most notably, we add a construct (\verb|Search|)
that searches for a natural-number value that makes an expression
evaluate to zero.  The value is substituted for a sole variable
(\verb|Var|) allowed by the language.  


\begin{figure}
\begin{verbatim}
data Val : Set where
 I : ℤ → Val
 B : 𝔹 → Val

data Expr : Set where
 Var : Expr
 Value : Val → Expr
 Add : Expr → Expr → Expr
 IsZero : Expr → Expr
 Cond : Expr → Expr → Expr → Expr
 Search : Expr → Expr
\end{verbatim}
\caption{Values and expressions}
\label{fig:syn}
\end{figure}

\section{The Evaluator}

The evaluator produces one of three possible results, as shown in
Figure~\ref{fig:res}.  It could produce a value (\verb|Value|),
failure (\verb|Fail|), or report that it ran out of fuel
(\verb|Unfinished|).  We are using a standard simple technique for
incorporating possibly diverging functions into a total programming
language (here, Agda), where the function takes in an extra
natural-number input, bounding the number of steps of computation.  If
this fuel parameter reaches zero and the computation is not complete,
then our evaluator will return \verb|Unfinished|.

\begin{figure}
\begin{verbatim}
data Result (V : Set) : Set where
 Value : V → Result V
 Fail : Result V
 Unfinished : Result V
\end{verbatim}
\caption{Results produced by the evaluator}
\label{fig:res}
\end{figure}

To reduce boilerplate in the evaluator, we define \verb|Result| as a monad, with the code in Figure~\ref{fig:resmon}.
Values are propagated by \verb|>>=|, but \verb|Fail| and \verb|Unfinished| terminate computation.  We make use of
Agda's approach to type classes using an \verb|instance| block, to declare that \verb|Result| is a monad.  

\begin{figure}
\begin{verbatim}
infixr 8 _>>=r_

_>>=r_ : ∀{A B : Set} → Result A → (A → Result B) → Result B
Value x >>=r r = r x 
Fail >>=r r' = Fail
Unfinished >>=r r' = Unfinished

returnr : ∀{A : Set} → A → Result A
returnr = Value

instance
  ResultMonad : Monad Result 
  ResultMonad = record { return = returnr ; _>>=_ = _>>=r_ }
\end{verbatim}
\caption{Monadic combinators for \texttt{Result}}
\label{fig:resmon}
\end{figure}

Finally, we define the evaluator with typing

\begin{verbatim}
eval : ℕ → Expr → ℕ → Result Val
\end{verbatim}

\noindent In call \verb|eval g e v|, the natural \verb|g| is the fuel, and
\verb|v| is the value for the sole variable.  We follow the proposed
structure of Garby et al., by having a helper function for each case
of evaluation.  So for evaluating an \verb|Add| expression, we have a
helper function \verb|add|.  Garby et al. give this helper the type
\verb|Value → Value → Value|.  This simple type hides some complexity,
because \verb|Error| is included in their type for values.  Here,
using the monadic structure of \verb|Result| we write:

\begin{verbatim}
add : Val → Val → Result Val
add (I x) (I y) = return (I (x +ℤ y))
add _ _ = Fail
\end{verbatim}

\noindent Our type \verb|Val| of values does not include errors (see
Figure~\ref{fig:syn}), so to return \verb|Fail|, the result type of
\verb|add| must use \verb|Result|.  The test for zero has just as
simple a helper function:

\begin{verbatim}
isZero : Val → Result Val
isZero (I x) = return (B (x =ℤ 0ℤ))
isZero _ = Fail
\end{verbatim}

\noindent It returns a boolean value if given an integer, and otherwise fails.


Because conditional is lazy in its then- and else-parts, we write the
following helper for \verb|Cond|:

\begin{verbatim}
cond : Val → Result Val → Result Val → Result Val
cond (B x) y z = if x then y else z
cond _ _ _ = Fail
\end{verbatim}

\noindent This takes in terms of type \verb|Result Val| for then- and else-parts, representing that computation
is not strict in those.

Deferring the code for \verb|search| for a moment, this brings us to the definition of \verb|eval| in Figure~\ref{fig:eval}.
We use \verb|do|-notation (supported by Agda) to use the monadic combinators for \verb|Result|.  In the case
for \verb|IsZero|, for example, we bind the result of the recursive call \verb|eval g e v| to \verb|r|, which
then has type \verb|Val|.  This value is given to the helper function \verb|isZero|, which returns a value
of type \verb|Result Val|, as required by the return type of \verb|eval|.

\begin{figure}
\begin{verbatim}
eval : ℕ → Expr → ℕ → Result Val
eval g Var v = return (I (toℤ v))
eval g (Value x) v = return x
eval g (Add e1 e2) v =
  do
    r1 ← eval g e1 v
    r2 ← eval g e2 v
    add r1 r2
eval g (IsZero e) v =
  do
    r ← eval g e v
    isZero r
eval g (Cond e1 e2 e3) v =
  do
    b ← eval g e1 v
    cond b (eval g e2 v) (eval g e3 v)
eval g (Search e) _ = search g (eval g e)
\end{verbatim}
\caption{The evaluator}
\label{fig:eval}
\end{figure}

\subsection{Implementing \texttt{Search}}

The missing piece of the evaluator is the code to process \verb|Search| expressions.  This
is given by helper function \verb|search| in Figure~\ref{fig:search}.  Some extension of~\citet{garby+25}
is required at this point, because on their methodology, the helper functions in the various cases
of \verb|eval| should be presented with values (\verb|Val|), or perhaps results of computation (\verb|Result Val|).
But for \verb|Search e|, we need to try evaluating \verb|e| repeatedly with different values for the variable.
So it seems that the helper function \verb|search| might need to deviate from the approach in~\citet{garby+25},
since it cannot take just a single value or result.

We solve this problem by calling \verb|search| with the partial
application \verb|eval g e|, which is missing the argument for the
value of the sole variable of \verb|e|.  The type of this partial
application is \verb|ℕ → Result Val|, which is what the code for
\verb|search| in Figure~\ref{fig:search} expects for its second
argument \verb|vv|.  The code for \verb|search| binds the value of
\verb|vv| on the current fuel \verb|g|, and then calls the
mutually recursive helper \verb|searchh|.  That function
checks to see if the value is integer zero.  If so, it returns
\verb|g| (in the \verb|Result| monad).  Otherwise, it either reports
\verb|Unfinished| if the fuel has been exhausted, or else recursively
continues the searches for a zero, decrementing the fuel from
\verb|suc g| to \verb|g| (in the last equation of
Figure~\ref{fig:search}).

\begin{figure}
\begin{verbatim}
mutual 
 search : ℕ → (ℕ → Result Val) → Result Val
 search g vv =
  do
   d ← vv g
   searchh g vv d 
 searchh : ℕ → (ℕ → Result Val) → Val → Result Val
 searchh g vv (B x) = Fail
 searchh g vv (I x) with x =ℤ 0ℤ
 searchh g vv (I x) | tt = return (I (toℤ g))
 searchh zero vv (I x) | ff = Unfinished
 searchh (suc g) vv (I x) | ff = search g vv
\end{verbatim}
\caption{The helper function for searching for a zero}
\label{fig:search}
\end{figure}

\section{The Syntax of Types}

Our aim is to define a type checker \verb|texp| satisfying the following version of the specification of~\citet{garby+25}:

\begin{verbatim}
texp e ≪ tresult (eval g e v)
\end{verbatim}

\noindent We will need a function \verb|tresult| to compute a type for
a value of type \verb|Result Val|, produced by \verb|eval|.
Differently from Garby et al., the relation $\ll$ that we use relates
two different kinds of types.  The type produced by \verb|texp| has
Agda type \verb|EType Data|, representing static types.  The type
produced by \verb|tresult| has Agda type \verb|RType|, for runtime
types.  The definitions are in Figure~\ref{fig:syntp}.  There is an
Agda type \verb|Data| for types common to both runtime and static.
Constructors \verb|DATA| (taking an argument of type \verb|Data|) and
\verb|FAIL| are common to both datatypes.  In the static case
(\verb|EType|), these constructors take a boolean argument indicating
whether the terms might diverge (\verb|True|) or certainly terminate
(\verb|False|).  For the situation where the evaluator runs out of
fuel and returns \verb|Unfinished|, we have runtime type
\verb|UNFINISHED|.  Statically, we will use a simple positivity
analysis to detect if the search for a zero cannot possibly succeed.
In that case, the static type is \verb|LOOP|.  Finally, static type
\verb|UNKNOWN| represents uncertainty, as generally inherent to type
checking.

\begin{figure}
\begin{verbatim}
data Data : Set where
  INT : Data
  BOOL : Data

data RType : Set where
  DATA : (c : Data) → RType
  UNFINISHED : RType
  FAIL : RType

data EType (X : Set) : Set where
  DATA : (b : 𝔹)(c : X) → EType X
  FAIL : 𝔹 → EType X
  UNKNOWN : EType X
  LOOP : EType X
\end{verbatim}
\caption{The syntax of runtime types (\texttt{RType}) and static types (\texttt{EType})}
\label{fig:syntp}
\end{figure}

The type constructor \verb|EType| is a monad, as shown by the code in
Figure~\ref{fig:etypemon}.  The definition of bind (\verb|>>=e|) for
\verb|EType| expresses that known failure (\verb|FAIL|), uncertainty
(\verb|UNKNOWN|), and known divergence (\verb|LOOP|) cut off type
checking.  This leaves the case where we are binding an \verb|EType A|
of the form \verb|DATA b t|.  This is when type checking has
determined that a term has \verb|Data| type \verb|t|, with boolean
\verb|b| indicating possibility of divergence.  In this case, we send
\verb|t| to the functional argument of bind.  But we use \verb|b|
to relax the \verb|EType| returned by \verb|f t|.  For if \verb|b| indicates
possible divergence, that possibility should be retained.  This is
implemented by the call \verb|b ⇓ f t|, using the function \verb|⇓| defined
in Figure~\ref{fig:down}.  If the type to be relaxed is of the form
\verb|DATA b' t| or \verb|FAIL b'|, then the relaxation uses \verb;b || b'; for the
possibility of divergence.  This means that if either the current type
or the input \verb|b| indicate that divergence is possible, then
so does the resulting type.  Otherwise, relaxation does not affect
the type.

\begin{figure}
\begin{verbatim}  
_>>=e_ : ∀{A B : Set} → EType A → (A → EType B) → EType B
DATA b t >>=e f = b ⇓ f t
FAIL b >>=e f = FAIL b
UNKNOWN >>=e f = UNKNOWN
LOOP >>=e f = LOOP

returne : ∀{A : Set} → A → EType A
returne x = DATA ff x

instance
  ETypeMonad : Monad EType
  ETypeMonad = record { return = returne ; _>>=_ = _>>=e_ }
\end{verbatim}  
\caption{Monadic combinators for \texttt{EType}}
\label{fig:etypemon}
\end{figure}

\begin{figure}
\begin{verbatim}  
_⇓_ : ∀{A : Set} → 𝔹 → EType A → EType A
b ⇓ DATA b' t = DATA (b || b') t
b ⇓ FAIL b' = FAIL (b || b')
b ⇓ t = t
\end{verbatim}  
\caption{Relaxing the possibility of divergence in a type}
\label{fig:down}
\end{figure}

\section{The Type Checker}

Typing of values and results is shown in Figure~\ref{fig:tpvr}.  Integer values (\verb|I|) have type \verb|INT|,
and boolean values (\verb|B|) \verb|BOOL|.  This is computed by \verb|tval|.  \verb|RType|s are computed for
\verb|Result Val|s by \verb|tresult|, in the obvious way: \verb|Fail| has type \verb|FAIL|, and \verb|Unfinished|
\verb|UNFINISHED|.  \verb|Value| results are mapped to \verb|DATA| types.  

\begin{figure}
\begin{verbatim}
tval : Val → Data
tval (I _) = INT
tval (B _) = BOOL

tresult : Result Val → RType
tresult (Value v) = DATA (tval v)
tresult Fail = FAIL 
tresult Unfinished = UNFINISHED
\end{verbatim}
\caption{Typing of results and values}
\label{fig:tpvr}
\end{figure}

The type checker \verb|texp| for expressions follows the structure
proposed by Garby et al.: there is a helper function for each form of
expression.  Here, since our type checker is monadic, the helper
functions are called with \verb|Data| types (i.e., values of type
\verb|Data|, from Figure~\ref{fig:syntp}), and produce \verb|Data|
types in the \verb|EType| monad.  For \verb|Add| and \verb|IsZero|
expressions, the helper functions are straightforward:

\begin{verbatim}
add' : Data → Data → EType Data
add' INT INT = return INT
add' _ _ = FAIL ff

isZero' : Data → EType Data
isZero' INT = return BOOL
isZero' _ = FAIL ff
\end{verbatim}

For conditionals (\verb|Cond|), the helper function needs to combine
the \verb|ETypes| obtained for the then- and else-parts.  We follow
Garby et al. in using the order-theoretic perspective, and defining

\begin{verbatim}
cond' : Data → EType Data → EType Data → EType Data
cond' BOOL t1 t2 = t1 ⊓ t2 
cond' _ _ _ = FAIL ff
\end{verbatim}

\noindent Here, \verb|t1 ⊓ t2| is the greatest lower bound of the \verb|ETypes|
\verb|t1| and \verb|t2|, in the partial order discussed next.

\subsection{An Ordering on Static Types}

Figure~\ref{fig:po} shows the partial order on \verb|ETypes|, from
which \verb|⊓| is derived.  At the bottom of the order we have
\verb|UNKNOWN|, and as we proceed upwards, we gain certainty.  The
types \verb|DATA tt INT|, \verb|DATA tt BOOL|, and \verb|FAIL tt| are
one level above \verb|UNKNOWN|.  The boolean \verb|tt| represents the
possibility of divergence.  So the type \verb|DATA tt INT| is for
terms which either diverge or else produce an integer value.  And the
type \verb|FAIL tt| is for terms which either diverge or exhibit a
runtime error, like trying to add an integer to a boolean.  These
types express less certainty than the similar types one level higher
with boolean \verb|ff|.  For example, \verb|DATA ff INT| is for terms
which are guaranteed to terminate with an integer value.  There is no
uncertainty about termination.  Finally, the type \verb|LOOP| is for
terms guaranteed to diverge.  This again reflects greater certitude
than the types like \verb|DATA tt INT|.


\begin{figure}
\begin{tikzpicture}
  \path (1,1) node(a){DATA tt INT};
  \path (4,1) node(b){DATA tt BOOL};
  \path (7,1) node(c){FAIL tt};
  \path (1,3) node(aa){DATA ff INT};
  \path (4,3) node(bb){DATA ff BOOL};
  \path (9,3) node(cc){FAIL ff};
  \path (4,-1) node(u){UNKNOWN};
  \path (7,3) node(l){LOOP};

  \draw[arrows = {-Stealth[length=5pt]}] (a.north) -- (aa.south);
  \draw[arrows = {-Stealth[length=5pt]}] (b.north) -- (bb.south);
  \draw[arrows = {-Stealth[length=5pt]}] (c.north) -- (cc.south);    
  \draw[arrows = {-Stealth[length=5pt]}] (u.north) -- (a.south);
  \draw[arrows = {-Stealth[length=5pt]}] (u.north) -- (b.south);
  \draw[arrows = {-Stealth[length=5pt]}] (u.north) -- (c.south);    
  \draw[arrows = {-Stealth[length=5pt]}] (a.north) -- (l.south);
  \draw[arrows = {-Stealth[length=5pt]}] (b.north) -- (l.south);
  \draw[arrows = {-Stealth[length=5pt]}] (c.north) -- (l.south);    
  
\end{tikzpicture}    
\caption{The partial order on types}
\label{fig:po}
\end{figure}

This ordering is implemented in the code for \verb|⊓|, shown in
Figure~\ref{fig:glb}.  The last two equations use the $\Downarrow$
function of Figure~\ref{fig:down} to relax the possibility of
divergence in a type.  For example, suppose we are combining
\verb|LOOP| with, say, \verb|FAIL ff|.  For conditionals, this would
represent having a then-part, say, which is known to diverge, and an
else-part which is known to fail.  Since we do not know statically
which branch will be taken, the best we can do for the type of the
whole conditional is \verb|FAIL tt|, representing the fact that the
term either diverges or else exhibits a runtime error.  This is what
will be produced by the call \verb|tt ⇓ (FAIL ff)|, and is indeed the
greatest lower bound of those types, in the ordering of
Figure~\ref{fig:po}.

\begin{figure}
\begin{verbatim}
_⊓_ : EType Data → EType Data → EType Data
t ⊓ UNKNOWN = UNKNOWN
UNKNOWN ⊓ t = UNKNOWN
DATA b c ⊓ DATA b' c' = if c =c c' then DATA (b || b') c else UNKNOWN
DATA _ _ ⊓ FAIL _ = UNKNOWN
FAIL _ ⊓ DATA _ _ = UNKNOWN
FAIL b ⊓ FAIL b' = FAIL (b || b')
t ⊓ LOOP = tt ⇓ t
LOOP ⊓ t = tt ⇓ t
\end{verbatim}
\caption{The greatest lower bound of \texttt{EType}s, with respect to the ordering depicted in Figure~\ref{fig:po}}
\label{fig:glb}
\end{figure}

\subsection{Positivity analysis}

The final helper function needed for the type checker \verb|texp| is
for \verb|Search| expressions.  We use this as an opportunity to
consider static detection of divergence, where the type checker
returns type \verb|LOOP|.  This is done here with a simple positivity
analysis for expressions (much like one that forms an initial example
in~\citet{garby+25}).  This analysis will be sound but on purpose, not
complete for solvability of integer expressions.  Small extensions of
the language (for example, including multiplication and multiple
variables) would make that problem undecidable, so it is in fact more
illustrative of the general situation to adopt a sound but incomplete
analysis.  If the linear integer polynomial is strictly positive,
then it has no solution.  So we test for this property.

We follow the pattern established already above.  We take a monadic
definition for sign, to account for the approximate nature of the
analysis.  This is shown in Figure~\ref{fig:sign}.  Signs are either
\verb|Known|, containing a boolean indicating either strictly positive
(\verb|tt|) or nonnegative (\verb|ff|), or else \verb|Unknown|.  We
make \verb|Sign| a monad by having \verb|Unknown| cut off the
computation.  Figure~\ref{fig:sexp} then gives the code for computing
signs for values, results, and expressions.  The code for \verb|sval|
matches on the form of integers in the library we are using.  They are
either \verb|mkℤ zero triv| for zero, \verb|mkℤ (suc n) tt| for a
strictly positive number, or \verb|mkℤ (suc n) ff| for a strictly
negative one.  We label the latter as \verb|Unknown|, because we
are deliberately keeping the analysis simple and approximate.

The code for \verb|sexp| returns \verb|Unknown| unless the expression
uses only \verb|Var|, \verb|Value|, and \verb|Add|.  We use the
monadic structure of \verb|Sign| to avoid explicity considering
\verb|Unknown| in the \verb|Add| case of \verb|sexp|.  There, if
either expression has sign \verb|tt| indicating strict positivity,
then so does the \verb|Add| expression.

\begin{figure}
\begin{verbatim}
data Sign (A : Set) : Set where
  Known : A → Sign A
  Unknown : Sign A

Pos : Sign 𝔹
Pos = Known tt

Nonneg : Sign 𝔹
Nonneg = Known ff

returns : ∀{A : Set} → A → Sign A
returns = Known

_>>=s_ : ∀{A B : Set} → Sign A → (A → Sign B) → Sign B
Known x >>=s f = f x
Unknown >>=s f = Unknown

instance
  SignMonad : Monad Sign 
  SignMonad = record { return = returns ; _>>=_ = _>>=s_ }
\end{verbatim}
\caption{A monadic definition of signs}
\label{fig:sign}
\end{figure}
 
\begin{figure}
\begin{verbatim}
sval : Val → Sign 𝔹
sval (I (mkℤ zero triv)) = Nonneg
sval (I (mkℤ (suc n) tt)) = Pos
sval (I (mkℤ (suc n) ff)) = Unknown
sval (B x) = Unknown

sresult : Result Val → Sign 𝔹
sresult (Value v) = sval v
sresult Fail = Unknown
sresult Loop = Unknown

sexp : Expr → Sign 𝔹
sexp Var = Nonneg
sexp (Value x) = sval x
sexp (Add e e') =
  do
    s ← sexp e
    s' ← sexp e'
    return (s || s')
sexp _ = Unknown
\end{verbatim}
\caption{Signs of values (\texttt{Val}), results (\texttt{Result Val}), and expressions (\texttt{Expr})}
\label{fig:sexp}
\end{figure}

\subsection{Helper Function for \texttt{Search}}

Using the type of signs from the previous section, Figure~\ref{fig:search}
gives the helper function for type-checking an expression of the form \verb|Search e|,
when \verb|e| has the given \verb|Data| type.  The first defining equation for \verb|search'|
says that it returns \verb|LOOP| if \verb|e| has type \verb|INT| and the sign is strictly
positive.  If the sign is anything else (second defining equation), the type is \verb|DATA tt INT|,
indicating that evaluation of \verb|Search e| will either diverge of produce an integer.  If
the type for \verb|e| is \verb|BOOL|, then the type is \verb|FAIL ff|, indicating that evaluation
of \verb|Search e| will definitely fail, because \verb|e| will evaluate to a boolean.  Of course,
this is just the helper function for \verb|Search e|.  It could happen that the \verb|EType| computed for \verb|e|
indicates possible or certain divergence, failure, or unknown behavior.  The monad takes care of all
those cases, in the code for \verb|texp|, which is shown next.

\begin{figure}
\begin{verbatim}
search' : Data → Sign 𝔹 → EType Data
search' INT (Known tt) = LOOP
search' INT _ = DATA tt INT
search' _ _ = FAIL ff
\end{verbatim}
\caption{Helper function for type-checking \texttt{Search} expressions}
\label{fig:search}
\end{figure}

\subsection{The Definition of the Type Checker}

Using the helper functions presented above, the code for the type
checker \verb|texp| is shown in Figure~\ref{fig:texp}.  We use the
monadic structure of \verb|EType| throughout, to abstract away
handling the various modes of typing, namely failure, certain
divergence, and possible divergence.  The case for \verb|Search|
expressions calls \verb|sexp| (Section~\ref{sec:sign}) to get the sign
of the expression \verb|e|, and then passes this to the helper
function \verb|search'|.  The monadic approach results in clean,
simple code, while still using helper functions as advocated by Garby
et al.  An even bigger payoff of the monadic approach will appear in
the next section, where we prove their soundness property, for this
richer setting.


\begin{figure}
\begin{verbatim}
texp : Expr → EType Data
texp Var = return INT
texp (Value v) = return (tval v)
texp (Add e1 e2) = 
 do
   t1 ← texp e1
   t2 ← texp e2
   add' t1 t2
texp (IsZero e) =
 do
   t ← texp e
   isZero' t
texp (Cond e1 e2 e3) =
 do
   t1 ← texp e1
   cond' t1 (texp e2) (texp e3)
texp (Search e) =
 do
   t ← texp e
   search' t (sexp e)
\end{verbatim}
\caption{The type checker \texttt{texp}}
\label{fig:texp}
\end{figure}

\section{Relating Typing and Evaluation}
\label{sec:snd}

In this section, we prove the specification proposed by Garby et al., that
the static type of an expression approximates the type of its runtime value.  We
formulate this soundness property as:

\begin{verbatim}
texp e ≪ tresult (eval g e v)
\end{verbatim}

\noindent where all the functions mentioned have been defined above,
and \verb|≪| is the relation between \verb|EType|s and \verb|RType|s
defined in Figure~\ref{fig:ll}.  Let us consider the various clauses
of the definition.  \verb|≪Unknown| says that the static type
\verb|UNKNOWN| approximates any runtime type \verb|T|.  This clause
expresses the same idea that Garby et al. proposed, as shown in the
definition of $\le$ in Figure~\ref{fig:orig}.  It allows the type
checker to signal uncertainty, whenever desired.  The clause
\verb|≪Data| says that if the static type is \verb|DATA b T|, then the
runtime type may be \verb|DATA T|.  The critical point here is that
the static type can indicate (via boolean \verb|b|) that a term might
diverge, and it is still acceptable if the runtime type indicates that
it actually did converge to a value of type \verb|T|.  A similar idea
is expressed in the clause \verb|≪Fail| for the \verb|FAIL| type.  The
clause \verb|≪Unfinished| expresses that any static type is acceptable
for a computation that has not yet finished.  This might seem
counterintuitive, because some static types indicate convergence.  In
fact, with our expression language, it cannot happen that a term is
statically judged to be terminating, but for some smaller values of
the fuel parameter to the interpreter evaluation produces
\verb|Unfinished|.  But the specification allows that, and it could
occur, if the type checker reported solvability of polynomials, not
just unsolvability.  So with our language, the only case where
\verb|≪Unfinished| can arise is when \verb|T| is \verb|LOOP|.  This
is, of course, a critical case, because it expresses the desired
soundness property when the type checker reports \verb|LOOP|: the
evaluator must always return \verb|Unfinished| in that situation,
to satisfy the specification.

\begin{figure}
\begin{verbatim}  
data _≪_ : EType Data → RType → Set where
  ≪Unknown : ∀ {T} → UNKNOWN ≪ T
  ≪Data : ∀{T}{b} → DATA b T ≪ DATA T
  ≪Fail : ∀{b} → FAIL b ≪ FAIL
  ≪Unfinished : ∀{T} → T ≪ UNFINISHED
\end{verbatim}  
\caption{The relation we prove between static and runtime types}
\label{fig:ll}
\end{figure}  

Several lemmas about the relation on types are listed in Figure~\ref{fig:lllem}.
The first says that relaxing the possibility of divergence in a static type
preserves the relation.  The next two express the basic order-theoretic
property that $\sqcap$ computes a lower bound.  These properties are also
used by Garby et al. for their relation.

\begin{figure}
\begin{verbatim}  
⇓mono : ∀{e : EType Data}{r : RType}{b : 𝔹} →
        e ≪ r →
        (b ⇓ e) ≪ r
⊓lb1 : ∀{a b : EType Data}{a' b' : RType} →
      a ≪ a' →
      b ≪ b' →
      a ⊓ b ≪ a'
⊓lb2 : ∀{a b : EType Data}{a' b' : RType} →
      a ≪ a' →
      b ≪ b' →
      a ⊓ b ≪ b'
\end{verbatim}  
\caption{Several lemmas about the relation between static and runtime types}
\label{fig:lllem}
\end{figure}  

\subsection{Monadic Structure of the Relation}

The relation $\ll$ of Figure~\ref{fig:ll} exhibits a monadic structure, which we use below
to abstract boilerplate from the proof of soundness.  The combinators corresponding
to monadic return and bind are shown in Figure~\ref{fig:llmon}.  The first, \verb|return≪|,
relates the values produced by the return combinators of the \verb|EType| and \verb|Result|
monads.  The second is more complex.  It aims to relate two binds:
\begin{enumerate}
\item \verb|a >>=e f| in the \verb|EType| monad, and
\item \verb|a' >>=r f'| in the \verb|Result| monad.
\end{enumerate}

\noindent These can be related if first, \verb|a ≪ tresult a'|.  Second,
we have a complex condition with a number of hypotheses:

\begin{itemize}
\item \verb|v : Val|
\item \verb|b : 𝔹|
\item \verb|a ≡ DATA b (tval v)|
\item \verb|a' ≡ Value v|
\item \verb|DATA b (tval v) ≪ DATA (tval v)|
\end{itemize}  

\noindent Providing all these, the monadic combinator then requires

\begin{itemize}
\item \verb|f (tval v) ≪ tresult (f' v)|
  \end{itemize}

\noindent The essence of this requirement (by the combinator, of its
caller) is to show that \verb|f| and \verb|f'| together preserve the
relation, for all \verb|v : Val|.  The extra premises are provided by
the combinator to its caller, to make this easier to show.  Providing
proofs that the terms \verb|a| and \verb|a'| referenced in the type of
the first argument to the combinator are equal to these specific
values (\verb|DATA b (tval v)| and \verb|Value v|) are justified by
the situation in which this second argument to the combinator will be
invoked; and also turn out to be quite helpful in using the
combinator.  The premise stating that these values are related by
$\ll$ is provided by the combinator just as a convenience in the
proof; it follows from the other premises and the requirement that
\verb|a ≪ tresult a'|.  Note that we cannot declare that this monadic
structure is an instance of \verb|Monad|, because the interfaces are
different: monads are not dependently typed, but these combinators
are.  So we cannot use \verb|do|-notation with these, but as we will
see next, we can still apply them directly with good effect.

\begin{figure}
\begin{verbatim}  
return≪ : ∀{v : Val} →
          returne (tval v) ≪ tresult (returnr v)
return≪{v} = ≪Data

_>>=≪_ : ∀{a : EType Data}{a' : Result Val} →
        {f : Data → EType Data}{f' : Val → Result Val} → 
         a ≪ tresult a' →
         (∀{v : Val}{b : 𝔹}
           {q : a ≡ DATA b (tval v)}
           {q' : a' ≡ Value v}
           {q'' : DATA b (tval v) ≪ DATA (tval v)} → 
           f (tval v) ≪ tresult (f' v)) → 
        (a >>=e f) ≪ tresult (a' >>=r f')
\end{verbatim}  
\caption{Monadic structure of $\ll$}
\label{fig:llmon}
\end{figure}  

\subsection{The Proof of Soundness: Base Cases}

We turn now to the proof of
\begin{verbatim}
texp-soundness : ∀{e : Expr}{g v : ℕ} →
                 texp e ≪ tresult (eval g e v)
\end{verbatim}
\noindent Let us start with the base cases:
\begin{verbatim}
texp-soundness {Value x} {g} {v} = return≪
texp-soundness {Var} {g} {v} = return≪{v = I (toℤ v)}
\end{verbatim}
\noindent They may both be proved using the \verb|return| of the monadic proof combinator.
In the \verb|Var| case, Agda cannot infer the value for implicit argument \verb|v|,
because in that case, the goal is
\begin{verbatim}
texp Var ≪ tresult (eval g Var v).
\end{verbatim}
\noindent which simplifies to just
\begin{verbatim}
DATA ff INT ≪ DATA INT
\end{verbatim}
\noindent So we supply the argument \verb|I (toℤ v)| explicitly to
\verb|return≪|.  We could also just use \verb|≪Data| (which is what
\verb|return≪| returns) for the \verb|Var| case.  Using \verb|return≪|
just helps illustrate the monadic nature of the proof.

\subsection{The Proof of Soundness: Arithmetic Cases}

Next, let us consider the case for \verb|IsZero|.  We must prove
\begin{verbatim}
texp (IsZero e) ≪ tresult (eval g (IsZero e) v)
\end{verbatim}
\noindent Unfolding the definitions of the type checker (\verb|texp|) and evaluator (\verb|eval|), this goal becomes

\begin{verbatim}
texp e >>=e isZero' ≪ tresult (eval g e v >>=r isZero)
\end{verbatim}

\noindent Delightfully, this matches the conclusion of our monadic proof combinator \verb|>>=≪|, which
is 
\begin{verbatim}
(a >>=e f) ≪ tresult (a' >>=r f')
\end{verbatim}
\noindent Even better, Agda can deduce that in order to apply that combinator, \verb|a| must be instantiated to
\verb|texp e|, \verb|f| to \verb|isZero'|, \verb|a'| to \verb|eval g e v|, and \verb|f'| to \verb|isZero|.
From the type of the combinator, this means we have two subgoals to prove, where the first is
\begin{verbatim}
texp e ≪ tresult (eval g e v)
\end{verbatim}
\noindent But this is just what the induction hypothesis (i.e., under
Curry-Howard, a recursive call to \verb|texp-soundness| produces for
\verb|e|).  So we need only prove the second premise of the
combinator.  The proof in Figure~\ref{fig:caseiszero} is sufficient;
it simply does not make use of the extra premises that the combinator
provides (about the identity of \verb|a|, etc.).  They are not needed
here.  The two cases of the proof, for the two constructors of
\verb|Val|, are easily proved from the defining clauses for $\ll$.
Those cases are:
\begin{enumerate}
\item \verb|isZero' (tval (I x)) ≪ tresult (isZero (I x))|
\item \verb|isZero' (tval (B x)) ≪ tresult (isZero (B x))|
\end{enumerate}
\noindent From the definitions of \verb|izZero'| and \verb|isZero|,
these simplify to 
\begin{enumerate}
\item \verb|DATA ff BOOL ≪ DATA BOOL|, because the static and runtime types of \verb|IsZero| expressions applied to integer values are both (forms of) the boolean type; and 
\item \verb|FAIL ff ≪ FAIL|, because \verb|IsZero| fails if applied to booleans.
\end{enumerate}
\noindent Notice that the monadic proof combinator has taken care of
all the error cases, and of propagating the possibility of divergence.
So here, it suffices to reason as if the subexpression \verb|e| is
terminating.

\begin{figure}
\begin{verbatim}
case-isZero : ∀{v : Val} →
              isZero' (tval v) ≪ tresult (isZero v)
case-isZero {I x} = ≪Data
case-isZero {B x} = ≪Fail
\end{verbatim}
\caption{Lemma for the \texttt{IsZero} case}
\label{fig:caseiszero}
\end{figure}

Using the monadic proof combinator and this three-line lemma \verb|case-isZero|, the case of \verb|texp-soundness| for \verb|IsZero|
is very easy:
\begin{verbatim}
texp-soundness {IsZero e} {g} {v} = 
  texp-soundness{e}{g}{v} >>=≪
  case-isZero  
\end{verbatim}
\noindent We just use the induction hypothesis (i.e., make a recursive
call) for \verb|e|, and bind the result to \verb|case-iszero| using
the proof combinator.


The case for \verb|Add| expressions works just as nicely.  We have this easily proven lemma,
where we relate data in the case of two integer values, and failures in all others:
\begin{verbatim}
case-add : ∀{v1 v2 : Val} →
           add' (tval v1) (tval v2) ≪ tresult (add v1 v2)
case-add {I _} {I _} = ≪Data
case-add {I _} {B _} = ≪Fail
case-add {B _} {I _} = ≪Fail
case-add {B _} {B _} = ≪Fail
\end{verbatim}
\noindent And the clause for \verb|texp-soundness| is again very simple, using the combinator:
\begin{verbatim}
texp-soundness {Add e1 e2} {g} {v} = 
  texp-soundness{e1}{g}{v} >>=≪
  texp-soundness{e2}{g}{v} >>=≪
  case-add
\end{verbatim}

\subsection{The Proof of Soundness: Conditionals}

Because evaluation of conditionals is lazy in the then- and else-parts, both the type checker and the evaluator
as we saw them above were structured somewhat differently.  The proof follows that structure.  The case
of \verb|texp-soundness| is:
\begin{verbatim}
texp-soundness {Cond e1 e2 e3} {g} {v} = 
  texp-soundness{e1}{g}{v} >>=≪ λ{v1} → 
  case-cond{v1} (texp-soundness{e2}{g}{v}) (texp-soundness{e3}{g}{v})
\end{verbatim}
\noindent We again use the proof-level bind \verb|>>=≪| with a
recursive call to \verb|texp-soundness| for the if-part of the
conditional.  The helper function then accepts the results of the
recursive calls for the then- and else-parts.  Agda has trouble
inferring some of the implicit arguments, so we handle them explicitly
using \verb|{v1}|.  Figure~\ref{fig:casecond} shows the definition of
the helper lemma \verb|case-cond|.  We pattern match on the value
\verb|v1| for the if-part, accepting proofs \verb|d1| and \verb|d2| of
the soundness property for the then- and else-parts of the
conditional.  If \verb|v1| is an integer value, then the goal
simplifies to 
\begin{verbatim}
FAIL ff ≪ FAIL
\end{verbatim}
\noindent This is proved by \verb|≪Fail|.  The other two cases are
for when \verb|v1| is \verb|B tt| or \verb|B ff|.  In those cases,
the goals simplify to
\begin{itemize}
\item \verb|t2 ⊓ t3 ≪ tresult r2|
\item \verb|t2 ⊓ t3 ≪ tresult r3|
\end{itemize}
\noindent We see here the \verb|⊓| operator, used in the definition of \verb|cond'| (the helper function used by 
the type checker \verb|texp|, for conditionals).  These goals are instances of the ordering properties \verb|⊓lb1|
and \verb|⊓lb2|, stated in Figure~\ref{fig:lllem} above.  So we prove those cases (in Figure~\ref{fig:casecond}) just using those lemmas.

\begin{figure}
\begin{verbatim}
case-cond : ∀{v1 : Val}{r2 r3 : Result Val}{t2 t3 : EType Data} →
            t2 ≪ tresult r2 →
            t3 ≪ tresult r3 →             
            cond' (tval v1) t2 t3 ≪ tresult (cond v1 r2 r3)
case-cond {I x} {r2} {r3} {t2} {t3} d1 d2 = ≪Fail
case-cond {B tt} {r2} {r3} {t2} {t3} d1 d2 = ⊓lb1 d1 d2
case-cond {B ff} {r2} {r3} {t2} {t3} d1 d2 = ⊓lb2 d1 d2
\end{verbatim}
\caption{Helper lemma for conditionals}
\label{fig:casecond}
\end{figure}

\subsection{Interlude: Soundness of the Positivity Analysis}

Before we see the proof of soundness for \verb|Search| expressions,
we need to consider soundness of the positivity analysis \verb|sexp|.
This is formulated again in the style of~\citet{garby+25}:
\begin{verbatim}
sexp-soundness : ∀{e : Expr}{g v : ℕ} →
                 sexp e ≪sign sresult (eval g e v)
\end{verbatim}
\noindent Here, the partial order is as shown in
Figure~\ref{fig:signll}, and it follows the same basic idea as the
ordering for types (of Figure~\ref{fig:ll}): lower in the ordering
corresponds to greater uncertainty.  The \verb|Unknown| type is a
minimal element, and \verb|Nonneg| is less than \verb|Pos| (as it
reflects the additional uncertainty of whether a term might evaluate
to zero or else be strictly positive).  The proof of
\verb|sexp-soundness| is simpler than the proof of soundness for
\verb|texp|, which we are in the middle of presenting.  For \verb|Add|
expressions, it uses a similar proof-level monadic bind operator,
shown in Figure~\ref{fig:bindllsign}.  As for typing, this operator is
for relating two monadic results:
\begin{verbatim}
(a >>=s f) ≪sign sresult (a' >>=r f')
\end{verbatim}
\noindent Also as for typing, it requires proofs of the relation, for the components of these
monadic results:
\begin{itemize}
\item \verb|a ≪sign sresult a'|
\item \verb|∀{b : 𝔹}{v : Val} → Known b ≪sign sval v → f b ≪sign sresult (f' v)|
\end{itemize}
\noindent The second condition states that if the sign of \verb|v| is known to be \verb|b|, then the
relation should hold of \verb|f b| and \verb|f' v|.  Using this combinator results in a very
simple case of \verb|sexp-soundness| for \verb|Add| expressions.


\begin{figure}
\begin{verbatim}  
data _≪sign_ : Sign 𝔹 → Sign 𝔹 → Set where
  ≪Unknown : ∀ {s} → Unknown ≪sign s
  ≪Refl : ∀{s} → s ≪sign s
  ≪Nonneg : Nonneg ≪sign Pos
\end{verbatim}  
\caption{The ordering on signs}
\label{fig:signll}
\end{figure}


\begin{figure}
\begin{verbatim}
_>>=≪sign_ : ∀{a : Sign 𝔹}{a' : Result Val} →
        {f : 𝔹 → Sign 𝔹}{f' : Val → Result Val} → 
         a ≪sign sresult a' →
         (∀{b : 𝔹}{v : Val}{q : a ≡ Known b}{q' : a' ≡ Value v} →
           Known b ≪sign sval v → 
           f b ≪sign sresult (f' v)) → 
        (a >>=s f) ≪sign sresult (a' >>=r f')
\end{verbatim}
\caption{Proof-level monadic bind, for proving soundness of \texttt{sexp}}
\label{fig:bindllsign}
\end{figure}


\subsection{The Proof of Soundness: Search}

Returning now to the proof of \verb|texp-soundness|, we have the last
case, namely for \verb|Search| expressions.  This is the most
involved, because both the type checker and evaluator are most
complicated for these expressions.  The type checker uses the
approximated positivity analysis \verb|sexp|.  The evaluator loops
through values for the variable, searching for zero, using mutually
recursive helper functions \verb|search| and \verb|searchh|.  Recall
that these functions use an input \verb|vv : ℕ → Result Val|, to avoid
having any direct dependence on expressions for the helper function.
We have (from Figure~\ref{fig:search})
\begin{itemize}
\item \verb|search g vv| searches for a value less than or equal to \verb|g| where
  \verb|vv| successfully returns zero.
\item \verb|searchh g vv u| checks if \verb|u| is zero, returning integer value \verb|g|, and continuing the search
  otherwise, if enough gas remains.
\end{itemize}
\noindent Not surprisingly, the proof needs mutually recursive lemmas to match this
structure.  These are stated in Figure~\ref{fig:lemsearch}



\begin{figure}
\begin{verbatim}
mutual
 case-search : ∀ {s : Sign 𝔹}{vv : ℕ → Result Val}{g : ℕ} →
               (∀{n : ℕ} → DATA tt INT ≪ tresult (vv n)) → 
               (∀{n : ℕ} → s ≪sign sresult (vv n)) →
               search' INT s ≪ tresult (search g vv)
 case-searchh : ∀ {u : Val}{s : Sign 𝔹}{vv : ℕ → Result Val}{g : ℕ}{b : 𝔹} →
               (∀{n : ℕ} → DATA b (tval u) ≪ tresult (vv n)) → 
               (∀{n : ℕ} → s ≪sign sresult (vv n)) →
               s ≪sign sval u → 
               search' (tval u) s ≪ tresult (searchh g vv u)
\end{verbatim}
\caption{Helper lemmas about \texttt{searchh} and \texttt{search}}
\label{fig:lemsearch}
\end{figure}

These lemmas require some additional hypotheses about
\verb|vv : ℕ → Result Val|.  First, with
\verb|s| the sign of the original expression \verb|e| (where \verb|e|
is not presented to these lemmas), both lemmas need the assumption
\begin{verbatim}
∀{n : ℕ} → s ≪sign sresult (vv n)
\end{verbatim}
\noindent This says the sign of the calls to \verb|vv| is approximated
by \verb|s|.  Where we invoke \verb|case-search| in the proof of
\verb|texp-soundness|, we will instantiate \verb|s| with \verb|sexp e|
and \verb|vv| with \verb|eval g e|.  So the required property about
the sign of \verb|vv| becomes
\begin{verbatim}
∀{n : ℕ} → sexp e ≪sign sresult (eval g e n)
\end{verbatim}
\noindent And this is exactly the conclusion of \verb|sexp-soundness|, which
we saw in the previous section.

The second requirement of \verb|case-search| is that
\begin{verbatim}
∀{n : ℕ} → DATA tt INT ≪ tresult (vv n)
\end{verbatim}
\noindent This says that \verb|vv| only produces values of data type \verb|INT|.
A similar requirement is imposed by \verb|case-searchh|.  These properties
are proved, in the body of \verb|texp-soundness|, by the induction
hypothesis, which applies to any value \verb|n| for the sole variable.
The proof of \verb|case-search| uses the proof-level bind, although Agda
needs more help in inferring the implicit arguments.  So the code does
not look as nice.  Still, the combinator abstracts the reasoning about
failure cases, and thus keeps the proofs shorter.

To conclude, the complete proof of soundness is listed in
Figure~\ref{fig:texpsnd}, omitting only local definitions of \verb|h|
and \verb|h'| in the final case.  These do some local rewriting with
the results of \verb|texp-soundness| and \verb|sexp-soundness|, which
is needed because Agda does not support rewriting inline, but only as
part of pattern matching.

\begin{figure}
\begin{verbatim}
texp-soundness : ∀{e : Expr}{g v : ℕ} →
                 texp e ≪ tresult (eval g e v)
texp-soundness {Value x} {g} {v} = return≪
texp-soundness {Var} {g} {v} = return≪{v = I (toℤ v)}
texp-soundness {Add e1 e2} {g} {v} = 
  texp-soundness{e1}{g}{v} >>=≪
  texp-soundness{e2}{g}{v} >>=≪
  case-add
texp-soundness {IsZero e} {g} {v} = 
  texp-soundness{e}{g}{v} >>=≪
  case-isZero  
texp-soundness {Cond e1 e2 e3} {g} {v} = 
  texp-soundness{e1}{g}{v} >>=≪ λ{v1} → 
  case-cond{v1}
    (texp-soundness{e2}{g}{v})
    (texp-soundness{e3}{g}{v})
texp-soundness {Search e} {g} {v} =  
  texp-soundness{e}{g}{g} >>=≪
   λ{u}{b}{q}{q'} → 
   case-searchh{u}{b = b}
      (λ{n} → h {u} q)
      (sexp-soundness{e})
      (h' q')
\end{verbatim}
\caption{Proof of soundness of the type checker with respect to the evaluator}
\label{fig:texpsnd}
\end{figure}

\section{Related Work}

\section{Conclusion and Future Work}

We have seen how to extend the inspiring approach of~\citet{garby+25}
to distinguish between errors and uncertainty.  The kinds of errors we
consider are divergence and finite failure.  We extend their example
expression language with a form of unbounded search, for which we use
a simple sound -- but intentionally incomplete -- divergence analysis.
The approach depends on a more complex partial ordering of types,
where total uncertainty (type \verb|UNKNOWN|) is minimal, and
progression in the ordering corresponds to increasing certainty.  We
used monads to manage the several different kinds of errors, both in
typing and evaluation.  To prove Garby et al.'s relational
specification -- that the static type approximates the runtime type of
the result of the interpreter -- we use a proof-level binary indexed
monad.  This factors out all the reasoning about errors, and results
in a concise and manageable proof of soundness, both for typing and
the positivity analysis (for divergence).  Using the standard
technique of a fuel-based interpreter, we are able to perform this
relational reasoning about possibly diverging evaluation.

Probably the most pressing avenue for future work is to extend the
approach of Garby et al. to higher-order functions.  Here, it seems
likely that the idea of relating static types of expressions with
runtime types of results of evaluation will break down.  This is
because in the presence of higher-order functions, computing the
runtime type for a functional value will no longer be simple or even,
for many type systems of interest, decidable.  Even if results are
restricted to be first-order, the soundness property will have to be
extended to higher order, to reason about intermediate steps of
computation.  For this, instead of relating static and runtime types,
one might be better off relating static types and results of
evaluation directly.  This might take the approach closer to usual
semantic approaches in Programming Languages, like step-indexed
logical relations~\cite{ahmed06}.

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper.bib}


\end{document}

